Parameter Recovery Validation

### 2025-06-13 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): adopt process duration with stick-breaking temporal prior
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

This validation study introduces **piecewise constant transcription rates** while maintaining analytical tractability and resolving critical dimensional analysis inconsistencies identified in the piecewise activation framework. The key innovation is a **dimensionally consistent parameterization** that eliminates parameter redundancy by fixing the reference transcription state, while maintaining biological interpretability through fold-change parameters.

@@

@@

This piecewise framework enables representation of diverse gene expression patterns through specific parameter combinations. Each pattern is defined by quantitative parameter ranges applied to **relative temporal parameters** that create distinct temporal dynamics when scaled by the global time parameter $T_M^*$.

**Parameter Framework**: The classification uses relative temporal parameters ($\tilde{t}^*_{0,\text{on}i}$, $\tilde{\delta}^*_i$) that are scaled by the global maximum time $T_M^*$ to produce absolute temporal parameters:

\begin{align}
t^*_{0,\text{on}i} &= T_M^* \times \tilde{t}^*_{0,\text{on}i} \quad \text{(absolute activation onset time)} \\
\delta^*_i &= T_M^* \times \tilde{\delta}^*_i \quad \text{(absolute activation duration)}
\end{align}

This hierarchical structure separates global temporal scaling from gene-specific temporal patterns, enabling consistent pattern classification across different experimental timescales.

@@

- $\tilde{t}^*_{0,\text{on}i} < 0.0$ (relative activation onset before observation window)
- $R_{\text{on},i} > 2.0$ (moderate to strong fold-change during activation)

**Biological interpretation**: These genes underwent activation before the experimental observation window. During the observed period, they exhibit decay-only dynamics as they return toward baseline expression levels. The negative relative onset time indicates that activation occurred in the "past" relative to the observation start.

@@

**Characteristics**: Genes that show complete activation-decay cycles within the observation window, exhibiting brief, pulse-like activation followed by return toward baseline.

@@

- $\tilde{t}^*_{0,\text{on}i} > 0.0$ (relative activation within observation window)
- $\tilde{t}^*_{0,\text{on}i} < 0.5$ (activation early enough to complete cycle within 50% of timeline)
- $\tilde{\delta}^*_i < 0.4$ (short activation duration, 40% of timeline or less)
- $R_{\text{on},i} > 2.0$ (sufficient fold-change to observe activation)

**Biological interpretation**: These genes activate early enough and for a short enough duration that the complete activation-decay cycle is observable within the experimental window. The constraints ensure that both the activation and subsequent decay phases are captured.

@@

**Characteristics**: Genes that become activated early and remain highly expressed throughout most of the observation period, showing net increase over the observation window.

@@

- $\tilde{t}^*_{0,\text{on}i} > 0.0$ (relative activation within observation window)
- $\tilde{t}^*_{0,\text{on}i} < 0.3$ (early activation onset within 30% of timeline)
- $\tilde{\delta}^*_i > 0.5$ (long activation duration, more than 50% of timeline)
- $R_{\text{on},i} > 2.0$ (strong fold-change during activation)

**Biological interpretation**: These genes activate early in the observation window and maintain high expression levels throughout most of the observed period. The long activation duration relative to the observation window means that decay back to baseline is either minimal or occurs beyond the observation period.

@@

The Normal prior for onset times enables modeling of **pre-activation scenarios** where genes are already activated when observation begins. This extension is mathematically valid and biologically realistic:

@@

1. **Phase determination remains consistent**: For any observed time $t^* \geq 0$, we can still determine which phase applies by comparing $t^*$ to $t^*_{0,\text{on}i}$ and $t^*_{0,\text{on}i} + \delta^*_i$

@@

3. **Biological interpretation is clear**: Negative onset simply means activation occurred before the observation window

@@

- **Phase 1 may be unobserved**: If $t^*_{0,\text{on}i} < 0$, no observed cells will be in Phase 1
- **Phase 2 starts immediately**: Cells at $t^* = 0$ may already be in the activated state
- **Duration constraints**: For pre-activation genes, $\delta^*_i > |t^*_{0,\text{on}i}|$ ensures activation extends into the observation window

@@

# Prior hyperparameters (updated for relative temporal parameterization)
priors = {
    'R_on': {'loc': 0.916, 'scale': 0.4},              # log(2.5), fold-change (LogNormal)
    'tilde_t_on_star': {'loc': 0.5, 'scale': 0.8},     # Normal(0.5, 0.8²), relative onset (allows negatives)
    'tilde_delta_star': {'loc': -0.8, 'scale': 0.45},  # log(0.45), relative duration (LogNormal)
}

# Calculate constraint feasibility for each pattern (updated to match script implementation)
patterns = {
    'pre_activation': [
        ('tilde_t_on_star', '<', 0.0),  # Activation before observation window
        ('R_on', '>', 2.0),             # Moderate to strong fold change
    ],
    'transient': [
        ('tilde_t_on_star', '>', 0.0),  # Activation within observation window
        ('tilde_t_on_star', '<', 0.5),  # Early enough to complete cycle (50% of timeline)
        ('tilde_delta_star', '<', 0.4), # Short pulse duration (40% of timeline)
        ('R_on', '>', 2.0),             # Sufficient fold change to observe
    ],
    'sustained': [
        ('tilde_t_on_star', '>', 0.0),  # Activation within observation window
        ('tilde_t_on_star', '<', 0.3),  # Early activation onset (30% of timeline)
        ('tilde_delta_star', '>', 0.5), # Long activation duration (50% of timeline)
        ('R_on', '>', 2.0),             # Strong fold change
    ]
}

@@

        if param == 'tilde_t_on_star':  # Normal distribution
            prob = calculate_normal_cdf_probability(
                prior['loc'], prior['scale'], threshold, upper=(op == '>')
            )
        else:  # LogNormal distribution
            prob = calculate_lognormal_cdf_probability(
                prior['loc'], prior['scale'], threshold, upper=(op == '>')
            )

@@

The probabilistic graphical model extends Study 2 by adding gene-specific piecewise activation parameters while maintaining the hierarchical latent time structure. For visual clarity, we use $\theta_i$ to represent the vector of non-temporal piecewise activation parameters $(\gamma^*_i, R_{\text{on},i})$ and $\Delta_i$ to represent the vector of relative temporal gene parameters $(\tilde{t}^*_{0,\text{on}i}, \tilde{\delta}^*_i)$. Note that $\alpha^*_{\text{off}i} = 1.0$ is fixed and not shown as a random variable.

@@

@@

@@

# time hierarchy edges
pgm.add_edge("T_max_star", "t_star_j", plot_params=edge_params)
pgm.add_edge("t_loc", "t_star_j", plot_params=edge_params)
pgm.add_edge("t_scale", "t_star_j", plot_params=edge_params)
# temporal scaling edge for gene parameters
pgm.add_edge("T_max_star", "Delta_i", plot_params=edge_params)

@@

The model combines the hierarchical latent time structure from Study 2 with piecewise constant transcription rates and steady-state initial conditions to resolve identifiability while enabling rich gene expression dynamics.

@@

- $\theta_i$ represents the vector of **non-temporal** piecewise activation parameters $(\gamma^*_i, R_{\text{on},i})$
- $\Delta_i$ represents the vector of **relative temporal** gene parameters $(\tilde{t}^*_{0,\text{on}i}, \tilde{\delta}^*_i)$
- The hyperparameters $\mu_{\theta}, \sigma_{\theta}^2$ represent the collection of hyperparameters for all components of $\theta_i$
- The hyperparameters $\mu_{\Delta}, \sigma_{\Delta}^2$ represent the collection of hyperparameters for all components of $\Delta_i$
- The absolute timing parameters $t^*_{0,\text{on}i} = T_{M}^* \times \tilde{t}^*_{0,\text{on}i}$ and $\delta^*_i = T_{M}^* \times \tilde{\delta}^*_i$ are computed deterministically
- $\alpha^*_{\text{off}i} = 1.0$ is fixed and not shown as a random variable
- $T_{M}^* \rightarrow \Delta_i$ represents the scaling relationship for all temporal parameters

**Detailed priors for constrained piecewise activation parameters:**

The priors are designed to encode biologically reasonable parameter ranges while eliminating dimensional redundancy and cleanly separating temporal from non-temporal parameters:

@@

**Relative temporal gene-specific parameters:**
\begin{align}
    \tilde{t}^*_{0,\text{on}i} &\sim \text{Normal}(0.5, 0.8^2) \quad \text{(relative activation onset time, allows negatives)} \\
    \tilde{\delta}^*_i &\sim \text{LogNormal}(\log(0.45), 0.45^2) \quad \text{(relative activation duration)}
\end{align}

**Absolute temporal parameters (deterministic transformations):**
\begin{align}
    t^*_{0,\text{on}i} &= T_{M}^* \times \tilde{t}^*_{0,\text{on}i} \quad \text{(absolute activation onset time)} \\
    \delta^*_i &= T_{M}^* \times \tilde{\delta}^*_i \quad \text{(absolute activation duration)}
\end{align}

**Pattern classification framework**: The three expression patterns are defined by constraints on the relative temporal parameters:

- **Pre-activation**: $\tilde{t}^*_{0,\text{on}i} < 0.0$ and $R_{\text{on},i} > 2.0$
- **Transient**: $0.0 < \tilde{t}^*_{0,\text{on}i} < 0.5$, $\tilde{\delta}^*_i < 0.4$, and $R_{\text{on},i} > 2.0$
- **Sustained**: $0.0 < \tilde{t}^*_{0,\text{on}i} < 0.3$, $\tilde{\delta}^*_i > 0.5$, and $R_{\text{on},i} > 2.0$

@@

@@

**Relative temporal parameters:**

- **$\tilde{t}^*_{0,\text{on}i} \sim \text{Normal}(0.5, 0.8^2)$**: Centers relative activation onset at 50% through the process timeline, with 95% between approximately -1.1 and +2.1. **Negative values** enable pre-activation scenarios where genes were already activated before observation began.

- **$\tilde{\delta}^*_i \sim \text{LogNormal}(\log(0.45), 0.45^2)$**: Centers relative activation duration at 45% of the process timeline, with 95% between ~0.18 and ~1.13. This enables both transient (short duration) and sustained (long duration) activation patterns.

@@

- **Pre-activation** ($\tilde{t}^*_{0,\text{on}i} < 0$): ~26.6% probability under the prior
- **Transient** ($0 < \tilde{t}^*_{0,\text{on}i} < 0.5$ and $\tilde{\delta}^*_i < 0.4$): Balanced probability for observable complete cycles
- **Sustained** ($0 < \tilde{t}^*_{0,\text{on}i} < 0.3$ and $\tilde{\delta}^*_i > 0.5$): Balanced probability for long-duration activation

**Temporal scaling transformation:**

- **Absolute onset**: $t^*_{0,\text{on}i} = T_{M}^* \times \tilde{t}^*_{0,\text{on}i}$ scales relative onset to absolute time
- **Absolute duration**: $\delta^*_i = T_{M}^* \times \tilde{\delta}^*_i$ scales relative duration to absolute time
- **Dimensional consistency**: All temporal comparisons now use consistently scaled absolute times

**Parameter interpretation benefits:**

- **Separation of concerns**: Non-temporal kinetics ($\theta_i$) vs temporal patterns ($\Delta_i$)
- **Scale invariance**: Relative temporal patterns are independent of absolute time scale
- **Biological realism**: Global time scale $T_{M}^*$ sets the pace for the entire process

@@

**Hierarchical priors for latent time coordinates (unchanged from Study 2):**
\begin{align}
    T_{M}^* &\sim \text{Gamma}(\alpha_{T}, \beta_{T}) \\
    t_{loc} &\sim \text{Gamma}(\alpha_{t_{loc}}, \beta_{t_{loc}}) \\
    t_{scl} &\sim \text{Gamma}(\alpha_{t_{scl}}, \beta_{t_{scl}}) \\
    \tilde{t}_j &\sim \text{Normal}(t_{loc}, t_{scl}^2) \quad \text{for } j \in \{1, \ldots, N\} \\
    t^*_j &= T_{M}^* \times \max(\tilde{t}_j, \epsilon) \quad \text{(clipped transformation)}
\end{align}

@@

The dimensionless concentrations are computed using the analytical solutions for each phase as derived above, with the appropriate phase determined by comparing $t^*_j$ to the gene-specific activation parameters $t^*_{0,\text{on}i}$ and $\delta^*_i$.

@@

where $\theta$ represents the complete parameter vector including all gene-specific parameters $(\gamma^*_i, R_{\text{on},i}, \tilde{t}^*_{0,\text{on}i}, \tilde{\delta}^*_i, U_{0i})$, hierarchical time parameters $(T_{M}^*, t_{loc}, t_{scl}, t^*_j)$, and cell-specific capture efficiencies $(\lambda_j)$.

@@

\begin{align}
p(\tilde{y}) &= \int p(\tilde{y} | \gamma^*_i, R_{\text{on},i}, \tilde{t}^*_{0,\text{on}i}, \tilde{\delta}^*_i, U_{0i}, T_{M}^*, t_{loc}, t_{scl}, t^*_j, \lambda_j) \times \nonumber \\
&\quad p(\gamma^*_i, R_{\text{on},i}, \tilde{t}^*_{0,\text{on}i}, \tilde{\delta}^*_i, U_{0i}, T_{M}^*, t_{loc}, t_{scl}, t^*_j, \lambda_j) \, d\theta \label{eq-prior-predictive-piecewise}
\end{align}

@@

\begin{align}
p(\tilde{y}) &= \int \int \int \int \int p(\tilde{y} | T_{M}^*, \theta_{\text{gene}}, \theta_{\text{cell}}, \theta_{\text{tech}}) \times \nonumber \\
&\quad p(T_{M}^*) \, p(\theta_{\text{gene}}) \, p(\theta_{\text{cell}}) \, p(\theta_{\text{tech}}) \, dT_{M}^* \, d\theta_{\text{gene}} \, d\theta_{\text{cell}} \, d\theta_{\text{tech}} \label{eq-hierarchical-integration}
\end{align}

@@

- $\theta_{\text{gene}} = \{\gamma^*_i, R_{\text{on},i}, \tilde{t}^*_{0,\text{on}i}, \tilde{\delta}^*_i, U_{0i}\}_{i=1}^G$ represents gene-specific parameters
- $\theta_{\text{cell}} = \{t_{loc}, t_{scl}, \tilde{t}_j, t^*_j\}_{j=1}^N$ represents cell-specific temporal parameters
- $\theta_{\text{tech}} = \{\lambda_j\}_{j=1}^N$ represents technical efficiency parameters

This decomposition reflects the model's assumption that gene-specific kinetics, population-level temporal structure, and technical factors contribute independently to the prior uncertainty.

@@

1. **Sample hierarchical time parameters**:
   \begin{align}
   {T_{M}^*}^{(s)} &\sim \text{Gamma}(\alpha_{T}, \beta_{T}), \quad {t_{loc}}^{(s)} \sim \text{Gamma}(\alpha_{t_{loc}}, \beta_{t_{loc}}), \quad {t_{scl}}^{(s)} \sim \text{Gamma}(\alpha_{t_{scl}}, \beta_{t_{scl}})
   \end{align}

2. **Sample cell-specific relative times**:
   \begin{align}
   {\tilde{t}_j}^{(s)} &\sim \text{Normal}({t_{loc}}^{(s)}, ({t_{scl}}^{(s)})^2) \quad \text{for } j \in \{1, \ldots, N\}
   \end{align}

3. **Transform to absolute times**:
   \begin{align}
   {t^{*}_j}^{(s)} &= {T_{M}^*}^{(s)} \times \max({\tilde{t}_j}^{(s)}, \epsilon)
   \end{align}

4. **Sample gene-specific parameters**:
   \begin{align}
   {\gamma^{*}_i}^{(s)} &\sim \text{LogNormal}(\log(1.0), 0.5^2) \\
   {R_{\text{on},i}}^{(s)} &\sim \text{LogNormal}(\log(2.5), 0.4^2) \\
   {\tilde{t}^*_{0,\text{on}i}}^{(s)} &\sim \text{Normal}(0.5, 0.8^2) \\
   {\tilde{\delta}^*_i}^{(s)} &\sim \text{LogNormal}(\log(0.45), 0.45^2) \\
   {U_{0i}}^{(s)} &\sim \text{LogNormal}(\mu_{U_0}, \sigma_{U_0}^2)
   \end{align}

5. **Transform relative temporal parameters**:
   \begin{align}
   {t^*_{0,\text{on}i}}^{(s)} &= {T_{M}^*}^{(s)} \times {\tilde{t}^*_{0,\text{on}i}}^{(s)}, \quad {\delta^*_i}^{(s)} = {T_{M}^*}^{(s)} \times {\tilde{\delta}^*_i}^{(s)}
   \end{align}

6. **Sample capture efficiencies**:
   \begin{align}
   {\lambda_j}^{(s)} &\sim \text{LogNormal}(\mu_{\lambda}, \sigma_{\lambda}^2)
   \end{align}

7. **Compute latent concentrations**:
   \begin{align}
   {u^{*}_{ij}}^{(s)}, {s^{*}_{ij}}^{(s)} &= f_{\text{piecewise}}({t^*_j}^{(s)}, {\gamma^*_i}^{(s)}, {R_{\text{on},i}}^{(s)}, {t^*_{0,\text{on}i}}^{(s)}, {\delta^*_i}^{(s)})
   \end{align}

8. **Sample observations**:
   \begin{align}
   {u_{ij}}^{(s)} &\sim \text{Poisson}({\lambda_j}^{(s)} \cdot {U_{0i}}^{(s)} \cdot {u^{*}_{ij}}^{(s)}) \\
   {s_{ij}}^{(s)} &\sim \text{Poisson}({\lambda_j}^{(s)} \cdot {U_{0i}}^{(s)} \cdot {s^{*}_{ij}}^{(s)})
   \end{align}

>>>>

This validation study introduces **piecewise constant transcription rates** while maintaining analytical tractability and resolving critical dimensional analysis inconsistencies identified in the piecewise activation framework. The key innovation is a **dimensionally consistent parameterization** that eliminates parameter redundancy by fixing the reference transcription state, while maintaining biological interpretability through fold-change parameters. Additionally, this study implements a **process duration with stick-breaking** temporal framework that eliminates scaling symmetry and achieves parameter identifiability while preserving all analytical and computational advantages.

@@

### Process Duration with Stick-Breaking Framework

This study implements a **process duration with stick-breaking** approach that eliminates the scaling symmetry present in hierarchical temporal scaling while maintaining all analytical and biological advantages. The key innovation is the use of **fixed anchor points** that break scaling symmetry and achieve parameter identifiability.

#### **Elimination of Scaling Symmetry**

The previous hierarchical scaling approach suffered from fundamental **scaling symmetry** that prevented parameter identifiability:

$$\mathcal{L}(T_M^*, \{t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\}) = \mathcal{L}(\lambda T_M^*, \{\lambda t^*_j\}, \{\lambda t^*_{0,\text{on}i}\}, \{\lambda \delta^*_i\})$$

for any $\lambda > 0$. This symmetry creates non-identifiable parameters and leads to catastrophic validation failures.

#### **Fixed Anchor Point Solution**

The stick-breaking approach eliminates scaling symmetry through **fixed anchor points**:

\begin{align}
t^*_1 &= 0 \quad \text{(process start, fixed)} \\
t^*_N &= T_{\max} \quad \text{(process end, sampled)} \\
t^*_{0,\text{on}i} &\sim \text{Normal}(\mu_{\text{on}}, \sigma_{\text{on}}^2) \quad \text{(independent gene timing)} \\
\delta^*_i &\sim \text{LogNormal}(\mu_{\delta}, \sigma_{\delta}^2) \quad \text{(independent gene duration)}
\end{align}

**Identifiability proof**: For any $\lambda \neq 1$:
$$\mathcal{L}(T_{\max}, \{t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\}) \neq \mathcal{L}(\lambda T_{\max}, \{\lambda t^*_j\}, \{t^*_{0,\text{on}i}\}, \{\delta^*_i\})$$

The fixed anchor $t^*_1 = 0$ cannot be scaled, and gene parameters are independent of $T_{\max}$, breaking the symmetry.

@@

This framework enables representation of diverse gene expression patterns through specific parameter combinations. Each pattern is defined by quantitative parameter ranges applied to **absolute temporal parameters** within the process duration framework.

**Parameter Framework**: The classification uses absolute temporal parameters that are independent of the process duration $T_{\max}$, enabling robust pattern identification across different experimental timescales.

@@

- $t^*_{0,\text{on}i} < 0.0$ (activation onset before process start)
- $R_{\text{on},i} > 2.0$ (moderate to strong fold-change during activation)

**Biological interpretation**: These genes underwent activation before the biological process under study began ($t^* = 0$). During the observed period, they exhibit decay-only dynamics as they return toward baseline expression levels. The negative onset time indicates that activation occurred before the process boundary.

@@

**Characteristics**: Genes that show complete activation-decay cycles within the process duration, exhibiting brief, pulse-like activation followed by return toward baseline.

@@

- $t^*_{0,\text{on}i} > 0.0$ (activation within process duration)
- $t^*_{0,\text{on}i} < 2.5$ (activation early enough to complete cycle within process)
- $\delta^*_i < 2.0$ (short activation duration relative to process)
- $R_{\text{on},i} > 2.0$ (sufficient fold-change to observe activation)

**Biological interpretation**: These genes activate early enough and for a short enough duration that the complete activation-decay cycle is observable within the process duration. The constraints ensure that both the activation and subsequent decay phases are captured within the temporal boundaries.

@@

**Characteristics**: Genes that become activated early and remain highly expressed throughout most of the process duration, showing net increase over the process timeline.

@@

- $t^*_{0,\text{on}i} > 0.0$ (activation within process duration)
- $t^*_{0,\text{on}i} < 1.5$ (early activation onset within process)
- $\delta^*_i > 2.5$ (long activation duration relative to process)
- $R_{\text{on},i} > 2.0$ (strong fold-change during activation)

**Biological interpretation**: These genes activate early in the process and maintain high expression levels throughout most of the process duration. The long activation duration means that decay back to baseline is either minimal or occurs beyond the process endpoint.

@@

The Normal prior for onset times enables modeling of **pre-activation scenarios** where genes are already activated when the biological process begins. This extension is mathematically valid and biologically realistic:

@@

1. **Phase determination remains consistent**: For any process time $t^* \geq 0$, we can still determine which phase applies by comparing $t^*$ to $t^*_{0,\text{on}i}$ and $t^*_{0,\text{on}i} + \delta^*_i$

@@

3. **Biological interpretation is clear**: Negative onset simply means activation occurred before the process start ($t^* = 0$)

@@

- **Phase 1 may be unobserved**: If $t^*_{0,\text{on}i} < 0$, no cells will be in Phase 1 during the process
- **Phase 2 starts immediately**: Cells at $t^* = 0$ may already be in the activated state
- **Duration constraints**: For pre-activation genes, $\delta^*_i > |t^*_{0,\text{on}i}|$ ensures activation extends into the process duration

@@

# Prior hyperparameters (updated for absolute temporal parameterization)
priors = {
    'R_on': {'loc': 0.916, 'scale': 0.4},              # log(2.5), fold-change (LogNormal)
    't_on_star': {'loc': 2.5, 'scale': 1.5},           # Normal(2.5, 1.5²), absolute onset (allows negatives)
    'delta_star': {'loc': 0.0, 'scale': 0.5},          # log(1.0), absolute duration (LogNormal)
}

# Calculate constraint feasibility for each pattern (updated for absolute parameterization)
patterns = {
    'pre_activation': [
        ('t_on_star', '<', 0.0),        # Activation before process start
        ('R_on', '>', 2.0),             # Moderate to strong fold change
    ],
    'transient': [
        ('t_on_star', '>', 0.0),        # Activation within process duration
        ('t_on_star', '<', 2.5),        # Early enough to complete cycle
        ('delta_star', '<', 2.0),       # Short pulse duration
        ('R_on', '>', 2.0),             # Sufficient fold change to observe
    ],
    'sustained': [
        ('t_on_star', '>', 0.0),        # Activation within process duration
        ('t_on_star', '<', 1.5),        # Early activation onset
        ('delta_star', '>', 2.5),       # Long activation duration
        ('R_on', '>', 2.0),             # Strong fold change
    ]
}

@@

        if param == 't_on_star':  # Normal distribution
            prob = calculate_normal_cdf_probability(
                prior['loc'], prior['scale'], threshold, upper=(op == '>')
            )
        else:  # LogNormal distribution
            prob = calculate_lognormal_cdf_probability(
                prior['loc'], prior['scale'], threshold, upper=(op == '>')
            )

@@

The probabilistic graphical model implements the process duration with stick-breaking framework while adding gene-specific piecewise activation parameters. For visual clarity, we use $\theta_i$ to represent the vector of non-temporal piecewise activation parameters $(\gamma^*_i, R_{\text{on},i})$ and $\Delta_i$ to represent the vector of absolute temporal gene parameters $(t^*_{0,\text{on}i}, \delta^*_i)$. Note that $\alpha^*_{\text{off}i} = 1.0$ is fixed and not shown as a random variable.

@@

# hyperparameters for process duration
pgm.add_node("alpha_T", r"$\alpha_{T}$", 5.5, 6, fixed=True)
pgm.add_node("beta_T", r"$\beta_{T}$", 6, 6, fixed=True)

# hyperparameters for lambda (entering from right edge of cell plate)
pgm.add_node("mu_lambda", r"$\mu_{\lambda}$", 7.5, 3.5, fixed=True)
pgm.add_node("sigma_lambda", r"$\sigma_{\lambda}^2$", 7.5, 2.5, fixed=True)

@@

# process duration parameter (replaces hierarchical time scaling)
pgm.add_node("T_max", r"$T_{\max}$", 5.5, 5)

# stick-breaking cell times and cell-specific parameters
pgm.add_node("t_star_j", r"$t^{\ast}_j$", 5.5, 4)
pgm.add_node("lambda_j", r"$\lambda_j$", 6.5, 3)

@@

# process duration edges
pgm.add_edge("alpha_T", "T_max", plot_params=edge_params)
pgm.add_edge("beta_T", "T_max", plot_params=edge_params)

# stick-breaking time edges
pgm.add_edge("T_max", "t_star_j", plot_params=edge_params)

@@

The model combines the process duration with stick-breaking framework with piecewise constant transcription rates and steady-state initial conditions to achieve parameter identifiability while enabling rich gene expression dynamics.

@@

- $\theta_i$ represents the vector of **non-temporal** piecewise activation parameters $(\gamma^*_i, R_{\text{on},i})$
- $\Delta_i$ represents the vector of **absolute temporal** gene parameters $(t^*_{0,\text{on}i}, \delta^*_i)$
- The hyperparameters $\mu_{\theta}, \sigma_{\theta}^2$ represent the collection of hyperparameters for all components of $\theta_i$
- The hyperparameters $\mu_{\Delta}, \sigma_{\Delta}^2$ represent the collection of hyperparameters for all components of $\Delta_i$
- $T_{\max}$ represents the total process duration with fixed anchor points at $t^*_1 = 0$ and $t^*_N = T_{\max}$
- $\alpha^*_{\text{off}i} = 1.0$ is fixed and not shown as a random variable
- Gene temporal parameters are sampled independently of $T_{\max}$, eliminating scaling symmetry

**Detailed priors for process duration with stick-breaking framework:**

The priors are designed to encode biologically reasonable parameter ranges while achieving parameter identifiability through fixed anchor points and independent gene timing:

**Process duration:**
\begin{align}
    T_{\max} &\sim \text{Gamma}(5.0, 1.0) \quad \text{(total process duration)}
\end{align}

@@

**Absolute temporal gene-specific parameters (independent of $T_{\max}$):**
\begin{align}
    t^*_{0,\text{on}i} &\sim \text{Normal}(2.5, 1.5^2) \quad \text{(absolute activation onset time)} \\
    \delta^*_i &\sim \text{LogNormal}(0.0, 0.5^2) \quad \text{(absolute activation duration)}
\end{align}

**Stick-breaking cell times:**
\begin{align}
    t^*_1 &= 0 \quad \text{(process start, fixed)} \\
    t^*_N &= T_{\max} \quad \text{(process end)} \\
    \xi_j &\sim \text{Beta}(1, N-j) \quad \text{for } j \in \{2, \ldots, N-1\} \\
    t^*_j &= t^*_{j-1} + \xi_{j-1}(T_{\max} - t^*_{j-1}) \quad \text{(recursive construction)}
\end{align}

**Pattern classification framework**: The three expression patterns are defined by constraints on the absolute temporal parameters:

- **Pre-activation**: $t^*_{0,\text{on}i} < 0.0$ and $R_{\text{on},i} > 2.0$
- **Transient**: $0.0 < t^*_{0,\text{on}i} < 2.5$, $\delta^*_i < 2.0$, and $R_{\text{on},i} > 2.0$
- **Sustained**: $0.0 < t^*_{0,\text{on}i} < 1.5$, $\delta^*_i > 2.5$, and $R_{\text{on},i} > 2.0$

@@

**Process duration:**

- **$T_{\max} \sim \text{Gamma}(5.0, 1.0)$**: Centers process duration at 5 dimensionless time units, providing sufficient range for exponential convergence while maintaining computational efficiency. The process has clear start ($t^* = 0$) and end ($t^* = T_{\max}$) boundaries.

@@

**Absolute temporal parameters:**

- **$t^*_{0,\text{on}i} \sim \text{Normal}(2.5, 1.5^2)$**: Centers activation onset at the middle of the expected process duration, with 95% between approximately -0.4 and +5.4. **Negative values** enable pre-activation scenarios where genes were already activated before the process began.

- **$\delta^*_i \sim \text{LogNormal}(0.0, 0.5^2)$**: Centers activation duration at 1 dimensionless time unit, with 95% between ~0.37 and ~2.7. This enables both transient (short duration) and sustained (long duration) activation patterns.

@@

- **Pre-activation** ($t^*_{0,\text{on}i} < 0$): ~4.8% probability under the prior
- **Transient** ($0 < t^*_{0,\text{on}i} < 2.5$ and $\delta^*_i < 2.0$): Balanced probability for observable complete cycles
- **Sustained** ($0 < t^*_{0,\text{on}i} < 1.5$ and $\delta^*_i > 2.5$): Balanced probability for long-duration activation

**Parameter independence benefits:**

- **Identifiability**: Gene timing parameters are independent of process duration, eliminating scaling symmetry
- **Biological realism**: Process boundaries ($0$ to $T_{\max}$) provide natural temporal reference frame
- **Cross-scale applicability**: Same parameter values work across different experimental timescales
- **Separation of concerns**: Process duration ($T_{\max}$) vs gene-specific timing (onset/duration) vs kinetics (degradation/fold-change)

@@

**Stick-breaking cell time coordinates:**

The cell times are sampled using a stick-breaking process that ensures proper ordering and eliminates scaling symmetry:

\begin{align}
    t^*_1 &= 0 \quad \text{(process start, fixed)} \\
    t^*_N &= T_{\max} \quad \text{(process end)} \\
    \xi_j &\sim \text{Beta}(1, N-j) \quad \text{for } j \in \{2, \ldots, N-1\} \\
    t^*_j &= t^*_{j-1} + \xi_{j-1}(T_{\max} - t^*_{j-1}) \quad \text{(recursive construction)}
\end{align}

This construction automatically ensures $0 = t^*_1 < t^*_2 < \cdots < t^*_N = T_{\max}$ and eliminates the need for clipping operations.

@@

The dimensionless concentrations are computed using the analytical solutions for each phase as derived above, with the appropriate phase determined by comparing $t^*_j$ to the gene-specific activation parameters $t^*_{0,\text{on}i}$ and $\delta^*_i$. The stick-breaking cell times provide the temporal coordinates for evaluating the piecewise analytical solutions.

@@

where $\theta$ represents the complete parameter vector including all gene-specific parameters $(\gamma^*_i, R_{\text{on},i}, t^*_{0,\text{on}i}, \delta^*_i, U_{0i})$, process duration parameter $(T_{\max})$, stick-breaking cell times $(t^*_j)$, and cell-specific capture efficiencies $(\lambda_j)$.

@@

\begin{align}
p(\tilde{y}) &= \int p(\tilde{y} | \gamma^*_i, R_{\text{on},i}, t^*_{0,\text{on}i}, \delta^*_i, U_{0i}, T_{\max}, t^*_j, \lambda_j) \times \nonumber \\
&\quad p(\gamma^*_i, R_{\text{on},i}, t^*_{0,\text{on}i}, \delta^*_i, U_{0i}, T_{\max}, t^*_j, \lambda_j) \, d\theta \label{eq-prior-predictive-piecewise}
\end{align}

@@

\begin{align}
p(\tilde{y}) &= \int \int \int \int p(\tilde{y} | T_{\max}, \theta_{\text{gene}}, \theta_{\text{cell}}, \theta_{\text{tech}}) \times \nonumber \\
&\quad p(T_{\max}) \, p(\theta_{\text{gene}}) \, p(\theta_{\text{cell}}) \, p(\theta_{\text{tech}}) \, dT_{\max} \, d\theta_{\text{gene}} \, d\theta_{\text{cell}} \, d\theta_{\text{tech}} \label{eq-hierarchical-integration}
\end{align}

@@

- $\theta_{\text{gene}} = \{\gamma^*_i, R_{\text{on},i}, t^*_{0,\text{on}i}, \delta^*_i, U_{0i}\}_{i=1}^G$ represents gene-specific parameters
- $\theta_{\text{cell}} = \{t^*_j\}_{j=1}^N$ represents stick-breaking cell times
- $\theta_{\text{tech}} = \{\lambda_j\}_{j=1}^N$ represents technical efficiency parameters

This decomposition reflects the model's assumption that gene-specific kinetics, process duration, stick-breaking cell times, and technical factors contribute independently to the prior uncertainty.

@@

1. **Sample process duration**:
   \begin{align}
   {T_{\max}}^{(s)} &\sim \text{Gamma}(5.0, 1.0)
   \end{align}

2. **Sample stick-breaking variables and construct cell times**:
   \begin{align}
   {\xi_j}^{(s)} &\sim \text{Beta}(1, N-j) \quad \text{for } j \in \{2, \ldots, N-1\} \\
   {t^*_1}^{(s)} &= 0 \\
   {t^*_N}^{(s)} &= {T_{\max}}^{(s)} \\
   {t^*_j}^{(s)} &= {t^*_{j-1}}^{(s)} + {\xi_{j-1}}^{(s)}({T_{\max}}^{(s)} - {t^*_{j-1}}^{(s)})
   \end{align}

3. **Sample gene-specific parameters**:
   \begin{align}
   {\gamma^{*}_i}^{(s)} &\sim \text{LogNormal}(\log(1.0), 0.5^2) \\
   {R_{\text{on},i}}^{(s)} &\sim \text{LogNormal}(\log(2.5), 0.4^2) \\
   {t^*_{0,\text{on}i}}^{(s)} &\sim \text{Normal}(2.5, 1.5^2) \\
   {\delta^*_i}^{(s)} &\sim \text{LogNormal}(0.0, 0.5^2) \\
   {U_{0i}}^{(s)} &\sim \text{LogNormal}(\mu_{U_0}, \sigma_{U_0}^2)
   \end{align}

4. **Sample capture efficiencies**:
   \begin{align}
   {\lambda_j}^{(s)} &\sim \text{LogNormal}(\mu_{\lambda}, \sigma_{\lambda}^2)
   \end{align}

5. **Compute latent concentrations**:
   \begin{align}
   {u^{*}_{ij}}^{(s)}, {s^{*}_{ij}}^{(s)} &= f_{\text{piecewise}}({t^*_j}^{(s)}, {\gamma^*_i}^{(s)}, {R_{\text{on},i}}^{(s)}, {t^*_{0,\text{on}i}}^{(s)}, {\delta^*_i}^{(s)})
   \end{align}

6. **Sample observations**:
   \begin{align}
   {u_{ij}}^{(s)} &\sim \text{Poisson}({\lambda_j}^{(s)} \cdot {U_{0i}}^{(s)} \cdot {u^{*}_{ij}}^{(s)}) \\
   {s_{ij}}^{(s)} &\sim \text{Poisson}({\lambda_j}^{(s)} \cdot {U_{0i}}^{(s)} \cdot {s^{*}_{ij}}^{(s)})
   \end{align}
